---
title: "Technical DifficulKEYS vs the Replicathon"
output: html_document
date: Replicathon 2020
author: Andrea Edwards, Otto Kailing, and Ana Kilgore
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # for code turnin
# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) # for presentation
```

## Introduction:

In this R Markdown document, we have documented our analysis (or attempted analysis) of two pharmacogenomic experiments and interpretation of their replicability. Our thought process was guided heavily by the provided materials, including this template and all the tutorials.

### Exploratory analysis of pharmacogenomic data

#### a. Basic attributes of the raw data

We began by exploring the raw Pharmaco Data file. In this dataset, there are 288 cell-lines and 15 drugs. The two studies, CCLE and GDSC, have 8 and 32 unique drug concentrations each respectively.This dataset has 6 different attributes; cell-line, drug type, dose ID, concentration, viability, and study.

```{r}
library(tidyverse)
## import data
pharmacoData <- readRDS(file.path("data", "rawPharmacoData.rds"))
```

```{r}
head(pharmacoData)
str(pharmacoData)
```

#### b. Viability Distribution

We used a histogram to view viability distribution, and compare it with the data descriptions we obtained earlier. Based on the values visible in the first lines of the dataset, the distribution shown by the histogram seems to match that of known viability values. 

```{r}
# creating histograms for the separate studies
pharmacoData %>%
  ggplot(aes(x = viability)) +
  geom_histogram(fill = "gray", color = "black") +
  facet_wrap(~ study) +
  ggtitle("Distributions of viability by study") 
range(pharmacoData$viability)
len <- length(pharmacoData$viability)
len
sum(pharmacoData$viability < 0) / len * 100
sum(pharmacoData$viability > 100) / len * 100
sum(pharmacoData$viability > 105) / len * 100
sum(pharmacoData$viability > 110) / len * 100
sum(pharmacoData$viability > 115) / len * 100
sum(pharmacoData$viability > 120) / len * 100

```

However, although the distribution fits, some of the outlying points do not make sense considering the attribute type. The term "viability" refers to the percentage of cells that survive upon drug exposure, so in theory, all viability points should fall within the expected range of 0-100.

<!-- This is wrong, was based on the wrong value. -->
<!-- According to the above calculations, data points have a range of -20.0 to 319.4919. From a total of 3,826,909 points, only 23 points fall under 0, while 15778 points are above 100. Overall, the outlying points make up only 0.4% of data, and are not statistically significant. This is most likely due to a technical issue. -->

According to the above calculations, data points have a range of -20.0 to 319.4919. From a total of 43427 points, 23 fall under 0, while 15778 points are above 100. Of the outlying points, 0.05% of points have viabilities below zero. This is so few that it's not an issue, and is most likely a technical error. 36.3% of points are over 100. This does not make sense given our understanding of viability. Even accounting for values close to 100, 5.79% of rows have a viability value over 120, which is well out of our expected range.

However, are both of the studies similar? Looking at the histograms again, it appears that the distribution of values over 100 is similar in CCLE and GDSC. So this does not represent a notable difference <em>between the two studies</em>.

#### c. Comparing Variation Distributions


```{r}
pharmacoData %>%
    ggplot(aes(x = viability, group = study, fill = study, color = study)) +
    geom_density(alpha = 1/4) +
    xlim(0, 170) +
    ggtitle("Distributions of viability scores by study")
```

The graph shown above shows a comparison of viability distributions across the two studies. Overlaying the two distributions, there do not seem to be any major variations in viability. 

```{r}
pharmacoData %>%
    ggplot(aes(x = viability, group = doseID, fill = doseID, color = doseID)) +
    geom_density(alpha = 1/4) +
    xlim(0, 170) +
    ggtitle("Distributions of viability scores by dose ID")
```

This distribution shows viability distributions across dosages 1-9. These distributions are not as consistent, at higher dosages viability becomes less consistent with lower doses. The high dose curves have higher viability at lower levels, and lower viability at higher levels.

#### Summarized Data Exploration

We also explored the Summarized Data Exploration data. The summarized Data file contains 6 columns representing the variables cell Line, drug, and IC50 and AUC for the CCLE study as well as for the GDSC study.

```{r}
summarizedData <- readRDS(file.path("data", "summarizedPharmacoData.rds"))
str(summarizedData)
```

### Using Correlation Measures to Assess Replicability of Drug Response Studies

#### Comparison of AUC across cell lines

```{r}
summarizedData %>%
    ggplot(aes(x = -log10(auc_GDSC), y = -log10(auc_CCLE))) +
    geom_point(alpha = 1/2, cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("-log10(GDSC AUC)") +
    ylab("-log10(CCLE AUC)") +
    ggtitle("AUC summaries of cell line response across studies")
```

Above we have representations of the various reactions cell Lines have with different drugs using AUC. To quantify these differences, we used Spearman correlations.

The calculating the correlation coefficients yielded fairly poor results, most of the values were below 0.5. The two drugs with the highest correlation coefficient were PD-0325901 (r = 0.607) and 17-AAG (r = 0.569). This suggests that across the two studies, there isn't particularly strong agreement.

```{r}
## with the tidyverse
drugCorrs <- summarizedData %>% 
    group_by(drug) %>%
    summarize(Spearman_auc = cor(-log10(auc_GDSC / 10^6), -log10(auc_CCLE / 10^6), method = "spearman"))
drugCorrs
```

#### Comparison of IC50 across cell lines

```{r}
summarizedData %>%
    ggplot(aes(x = -log10(ic50_GDSC), y = -log10(ic50_CCLE))) +
    geom_point(alpha = 1/2, cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("-log10(GDSC IC50)") +
    ylab("-log10(CCLE IC50)") +
    ggtitle("IC50 summaries of cell line response across studies")
```

```{r}
## with the tidyverse
drugCorrs <- summarizedData %>% 
    group_by(drug) %>%
    summarize(Pearson_ic50  = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "pearson"))
drugCorrs
```

We tried the same process with IC50, hoping to find greater agreement between the studies. However, IC50 correlations performed even worse than AUC's. Notably, some of the same drugs had the high correlation coefficients, PD-0325901 (r = 0.625) had the highest closely followed by 7 Nilotinib (r = 0.611). 17-AAG (r = 0.543), one of the highest in the AUC correlation calculations, had the third highest correlation coefficient in the IC50 calculations. The most consistent drug across both trials was 6 lapatinib, with a coefficient of r = 0.427 in the IC50 calculations vs r = 0.429 in the AUC calculations.

```{r}
## with the tidyverse
drugCorrs <- summarizedData %>% 
    group_by(drug) %>%
    summarize(Pearson_ic50  = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "pearson"),
              Spearman_auc = cor(-log10(auc_GDSC / 10^6), -log10(auc_CCLE / 10^6), method = "spearman"))
## with the tidyverse
drugCorrs <- gather(drugCorrs, measure, correlation, -drug)
drugCorrs %>%
    ggplot(aes(x = drug, y = correlation, fill = measure, group = measure)) +
    geom_bar(stat = "identity", position = position_dodge(), colour = "black") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_grey() +
    ylim(0, 1) + 
    ggtitle("Correlation of cell line IC50 summaries between studies for each drug")
```

### Identifying biological factors that influence replicability of pharmacogenomic studies

#### Analyzing Resistance and Sensitivity in Cell Lines
There seem to be some cell lines that seem to be consistently resistant, but the exclusion of these lines wouldn't increase replicability. The graph below shows the greatest correlation between studies of points close to (0,0) on the graph, which when looking at AUC values indicates higher resistance.

```{r}
AUC_study1 <- rbeta(200, 1, 5)
AUC_study2 <- rbeta(200, 1, 5)
resistant <- data.frame(AUC_study1, AUC_study2)
ggplot(resistant, aes(y = AUC_study2, x = AUC_study1)) +
    geom_point() +
    xlim(0, 1) +
    ylim(0, 1) +
    ggtitle("Simulated AUC of resistant cell lines")
```

The cell lines can be manipulated to show more resistant and more sensitive values. Looking at the graph below, our group determined an appropriate cutoff for resistance would be the vertical and horizontal assymptotes at (0, 25) and (25, 0).

```{r}
AUC_study1 <- c(rbeta(100, 1, 5), rbeta(100, 4, 2))
AUC_study2 <- c(rbeta(100, 1, 5), rbeta(100, 4, 2))
resistant <- data.frame(AUC_study1, AUC_study2, 
                        CellLine = c(rep("Resistant", 100), rep("Sensitive", 100)))
ggplot(resistant, aes(y = AUC_study2, x = AUC_study1, color = CellLine)) +
    geom_point() +
    xlim(0, 1) +
    ylim(0, 1) +
    ggtitle("Simulated AUC with half sensitive and half resistant cell lines")
```

#### Applying Resistance/Sensitivity principles to drugs

Using the principles we explored earlier, we can apply resistance and sensitivity principles to individual drugs, and determine the similarity of findings between the two studies.

```{r}
summarizedData <- summarizedData %>% 
              mutate(cutoff = ifelse(drug == "paclitaxel", 0.4, 0.1),
                     sensitivity_GDSC = factor(ifelse( auc_GDSC < cutoff, "Resistant", "Sensitive")), 
                     sensitivity_CCLE = factor(ifelse( auc_CCLE < cutoff, "Resistant", "Sensitive"))) 
table("GDSC" = summarizedData$sensitivity_GDSC,
      "CCLE" = summarizedData$sensitivity_CCLE)
```


```{r}
summarizedData <-
    summarizedData %>% 
    mutate(category = paste(sensitivity_GDSC, sensitivity_CCLE),
           category = fct_recode(category,
                                 "Both Resistant" = "Resistant Resistant",
                                 "Both Sensitive" = "Sensitive Sensitive",
                                 "GDSC Resistant / CCLE Sensitive" = "Resistant Sensitive",
                                 "GDSC Sensitive / CCLE Resistant" = "Sensitive Resistant"))
summarizedData %>%
    subset(drug == "PLX4720") %>%
    ggplot(aes(x = auc_GDSC, y = auc_CCLE, colour = category)) +
    geom_point() + 
    xlab("GDSC AUC") +
    ylab("CCLE AUC") +
    geom_hline(aes(yintercept = cutoff), colour="grey", alpha=0.75, lty=2) +
    geom_vline(aes(xintercept = cutoff), colour="grey", alpha=0.75, lty=2) +
    ggtitle("Drug PLX4720")
```

The above graph represents agreement (pink and purple) and contradictory information (green and blue) about resistance and sensitivity to drugs between the two studies. 

```{r}
ggplot(summarizedData, aes(x = auc_GDSC, y = auc_CCLE, colour = category)) +
    geom_point(cex = 0.5) + 
    facet_wrap(~ drug) +
    xlab("GDSC AUC") +
    ylab("CCLE AUC") +
    geom_hline(aes(yintercept = cutoff), colour = "grey", alpha = 0.75, lty = 2) +
    geom_vline(aes(xintercept = cutoff), colour = "grey", alpha = 0.75, lty = 2) +
    ggtitle("Cell line sensitivity classifications between studies")
```

Looking at the collection of scatterplots for individual drugs, there is obviously variation in the overlap of the two studies. Drugs like 17-AAG, Nilotinib, and PD-0325901 (which all, in the previous step, displayed had high correlation coefficient values) appear consist of a majority of pink/purple dots. To quantify the relationship between the studies, we can compute the Matthews correlation coefficient as shown below.

```{r}
## creation of a Matthews correlation coefficient
mcc <- function (study1, study2) {
    BS <- sum(study1 == "Sensitive" & study2 == "Sensitive") 
    BR <- sum(study1 == "Resistant" & study2 == "Resistant") 
    SR <- sum(study1 == "Sensitive" & study2 == "Resistant") 
    RS <- sum(study1 == "Resistant" & study2 == "Sensitive") 
    
    if (BS+SR == 0 | BS+RS == 0 | BR+SR == 0 |  BR+RS ==0){
        mcc <- ((BS*BR)-(SR*RS)) 
    }else{
        mcc <- ((BS*BR)-(SR*RS)) / sqrt(exp((log(BS+SR)+log(BS+RS)+log(BR+SR)+log(BR+RS))))
    }
    return(mcc)
}
drugCorrs <- summarizedData %>% 
    group_by(drug) %>% 
    summarise(matthews_corr = mcc(sensitivity_GDSC, sensitivity_CCLE))
drugCorrs
```

The Matthews Correlation Coefficients calculated indicate low levels of correlation between the two studies, none of the drugs have an r value greater than 0.5. Visually, drugs with higher MCC's appear to contain more pink/purple dots, indicating agreement. 5 Erlotinib (r = 0.432) and 6 lapatinib (r = 0.424) had the highest MCC's.

```{r}
ggplot(drugCorrs, aes(x = drug, y = matthews_corr)) +
    geom_bar(stat = "identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab("Drug") +
    ylab("Matthews Correlation Coefficient")
```


## Overall Findings:

Based on the low values for Spearman correlation between the GDSC and CCLE studies' AUC measurements, there does not seem to be a great deal of concurrence between studies.

We noticed that the cell lines with lower AUC values, aka higher resistance values, had a greater correlation across the two studies. However, even these were not exceptionally correlated. Correlation coefficients were higher when using AUC to compare the studies, rather than IC50. This also signifies that more resistant cell lines would be better to use in further, similar studies.

With regard to the biological factors influencing replicability: AUC values are more correlated (between the studies) the lower they get, which further reinforces our previous observations. 

We chose the low cutoffs for resistance vs. sensitivity that were originally in the tutorial, for consistency and so we could check our points. Plotting sensitivity to each drug shows varied levels between correspondence between studies for each drug, further backing our findings that replicability between the studies is spotty.
